---
title: "analysis"
author: "Patrick"
date: "2023-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(RSocrata)
library(tigris)
library(sf)
library(leaflet)
library(htmltools)
library(gt)
library(janitor)
library(lubridate)

```

get all my data from open data
```{r}
scooter_trips <- read.socrata("https://data.cityofchicago.org/resource/2i5w-ykuw.csv")

scooter_trips_tract <- read.socrata("https://data.cityofchicago.org/resource/cini-k95q.csv")

equity_priority_area <- read.socrata("https://data.cityofchicago.org/resource/99tm-6k6i.csv") %>% 
  st_as_sf(wkt = "polygon") %>% 
  st_union()

community_areas <- read.socrata("https://data.cityofchicago.org/resource/igwz-8jzy.csv") %>% 
  st_as_sf(wkt = "the_geom")

bike_lanes <- read.socrata("https://data.cityofchicago.org/resource/hvv9-38ut.csv")%>% 
  st_as_sf(wkt = "the_geom")

streets <- read.socrata("https://data.cityofchicago.org/resource/pr57-gg9e.csv")%>% 
  st_as_sf(wkt = "the_geom")
```

Number of scooter trips over time
```{r}

scooter_trips_clean <- scooter_trips %>% 
  mutate(date = floor_date(start_time, "month"))

scooter_trips_sum <- scooter_trips_clean %>% 
  group_by(date) %>% 
  summarize(trips = n())


```

rides over the year
```{r}
ggplot(scooter_trips_sum)+
  geom_line(mapping = aes(x = as.Date(date), y = trips))+
  scale_x_date(date_labels = "%m-%Y")+
  scale_y_continuous(label = scales::comma)
```

census tracts shapes
```{r}

tracts_sf <- tracts(
  state = "IL",
  county = "Cook",
  cb = T
)

```
creating sf
```{r}

scooter_trips_sf <- scooter_trips %>% 
  filter(!is.na(start_centroid_longitude)) %>% 
  st_as_sf(coords = c("start_centroid_longitude", "start_centroid_latitude"), remove = F)

scooter_trips_equity <- scooter_trips_sf %>% 
  mutate(equity_zone = lengths(st_intersects(.,equity_priority_area)))

scooter_trips_equity %>% as.data.frame() %>% 
  group_by(equity_zone == 1, vendor == "Lyft") %>% 
  summarize(n = n())
```
create cloropleth data sets. going to use community area because the tract data doesn't have it available by vendor
```{r}

scooter_trips_tract_sf <- scooter_trips_tract %>% 
  mutate(GEOID = as.character(start_census_tract)) %>% 
  left_join(tracts_sf, by = "GEOID") %>% 
  st_as_sf() %>% 
  st_set_crs(st_crs(scooter_trips_sf))

scooter_tract_sum <- scooter_trips_tract_sf %>% 
  group_by(GEOID) %>% 
  summarize(trips = sum(trip_count, na.rm = T)) %>% 
  filter(!is.na(GEOID))

scooter_communityarea <- community_areas %>% 
  left_join(scooter_trips, by = c("area_num_1" = "start_community_area_number")) %>% 
  #filter(vendor != "Lyft") %>% 
  group_by(community) %>% 
  summarize(rides = n(),
            no_lyft_rides = sum(vendor != "Lyft"))

```
test map
```{r}
ggplot(scooter_tract_sum)+
  geom_sf(mapping = aes(fill = trips))
```


Map of scooter trips
```{r}

tag.map.title <- tags$style(HTML("
  .leaflet-control.map-title {
    position: fixed !important;
    padding: 10px;
    background: rgba(255,255,255,0.8);
    color: black;
    border-radius: 5px;
  }
  .title {
    padding: 0px;
    margin: 0px;
    font-size: 20px;
    margin-block: 0px;
    font-weight: 800;
  }
  span {
    font-weight: 800;
    color: green;
  }
  p {
    font-size: 15px;
    margin: 0px;
    padding: 10px;
    margin-block: 0px;
  }
  img {
  max-width: 428px
  }
  @media (max-width:475px) {
  img {max-width:280px}
  .title {font-size: 13px}
  p {font-size: 9px}
  }
  
"))

title <- tags$div(
  tag.map.title, HTML("<img src = https://patrickspauster.com/images/streetsblog-logo.png><p class = 'title'>Chicago's E-Scooter Rides</p><p>Rides from May to December 2022<br><span>Select a neighborhood</span> to view the number of rides</p>")
)   

pal <- colorNumeric(
  palette = "Greens",
  domain = scooter_communityarea$rides)

no_lyft_pal <- colorNumeric(
  palette = "Greens",
  domain = scooter_communityarea$no_lyft_rides)

map <- leaflet(scooter_communityarea, options = leafletOptions(zoomControl = FALSE)) %>% 
  addProviderTiles(provider = "CartoDB.Positron") %>% 
  addPolygons(color = ~pal(rides),
              opacity = 0.9,
              weight = 0.5,
              group = "Total") %>% 
  addPolygons(color = ~no_lyft_pal(no_lyft_rides),
              opacity = 0.9,
              weight = 0.5,
              group = "Dockless") %>% 
  addPolygons(data = equity_priority_area,
              color = "red",
              weight = 0.5,
              opacity = 0.5) %>% 
  addLegend("bottomright", pal = pal, values = ~rides,
    title = "Total",
    opacity = 1,
    group = "Total"
  ) %>% 
  addLegend("bottomright", pal = no_lyft_pal, values = ~no_lyft_rides,
    title = "Dockless",
    opacity = 1,
    group = "Dockless"
  ) %>% 
  #addLegend("bottomleft", colors = "red", labels = "equity zone") %>% 
  addControl(title,
             className = "map-title",
             position = "topleft") %>% 
  addLayersControl(
    baseGroups = c("Total", "Dockless"),
    options = layersControlOptions(collapsed = F)
  ) %>% 
  htmlwidgets::onRender("function(el, x) {
        L.control.zoom({ position: 'bottomleft' }).addTo(this)
    }")%>%
  htmlwidgets::onRender("
    function(el, x) {
      var updateLegend = function () {
          var selectedGroup = document.querySelectorAll('input:checked')[0].nextSibling.innerText.substr(1);

          document.querySelectorAll('.legend').forEach(a => a.hidden=true);
          document.querySelectorAll('.legend').forEach(l => {
            if (l.children[0].children[0].innerText == selectedGroup) l.hidden=false;
          });
      };
      updateLegend();
      this.on('baselayerchange', e => updateLegend());
     }")

#add transparent equity zone
#change picture
#add popups

map

mapview::mapshot(map, "map.html")
```




Differences in vendors?
Lyft scooters are all docked - need to break these out because they are all downtown and not in the equity area.
```{r}

count(scooter_trips_equity %>% as.data.frame() %>% group_by(vendor), equity_zone) %>% 
  pivot_wider(names_from = "equity_zone", values_from = "n") %>% 
  adorn_totals("col") %>% 
  mutate(percent_equity_zone = scales::percent(`1`/(`0`+`1`)))
  
count(scooter_trips_clean, vendor) %>% 
  adorn_totals() %>% 
  adorn_percentages("col")
```
Lyft only
```{r}
lyft_only <- scooter_trips_equity %>%
  filter(vendor == "Lyft") %>% 
  group_by(start_centroid_latitude, start_centroid_longitude) %>% 
  summarize(count = n()) %>% 
  arrange(count)

divvy_stations <- read.socrata("https://data.cityofchicago.org/resource/bk89-9dk7.json") %>% 
  st_as_sf(coords = c("location.longitude", "location.latitude"))

leaflet(lyft_only) %>% 
  addProviderTiles(provider = "CartoDB.Positron") %>% 
  addCircleMarkers(color = "pink",
              fillOpacity = 0.9,
              radius = ~sqrt(count)/10) %>% 
  addMarkers(lng = lyft_only$start_centroid_longitude,
             lat = lyft_only$start_centroid_latitude)
```
trips but desitnation - just verifying origin and destination doesn't make a big difference
```{r}

end_scooter_trips_sf <- scooter_trips %>% 
  filter(!is.na(end_centroid_longitude)) %>% 
  st_as_sf(coords = c("end_centroid_longitude", "end_centroid_latitude"), remove = F)

end_scooter_trips_equity <- end_scooter_trips_sf %>% 
  mutate(equity_zone = lengths(st_intersects(.,equity_priority_area)))

count(end_scooter_trips_equity %>% as.data.frame(), equity_zone == 1) %>% 
  janitor::adorn_percentages("col")

```
Math for ridership from the pilots - pulled from the pdfs
```{r}
twenty20 <- 540035/122
twenty19 <- 664975/122
program <- 6300

```
45% more time, 50% more distance
```{r}
scooter_trips_equity %>% 
  as.data.frame() %>% 
  group_by(equity_zone, vendor == "Lyft") %>% 
  summarize(mean(trip_distance, na.rm = T),
            mean(trip_duration, na.rm = T))
```
bike lanes and equity zone stats
```{r}

lanes_equity <- st_intersection(bike_lanes, equity_priority_area, )
  
lanes_no_equity <- st_difference(bike_lanes, equity_priority_area)
  
length_eq <- sum(st_length(lanes_equity))

length_no_eq <- sum(st_length(lanes_no_equity))
  
length_no_eq/length_eq

length_eq/(length_eq+length_no_eq)

```
streets proportion stats 
```{r}

streets_equity <- st_intersection(streets, equity_priority_area)
  
streets_no_equity <- st_difference(streets, equity_priority_area)
  
slength_eq <- sum(st_length(streets_equity))

slength_no_eq <- sum(st_length(streets_no_equity))
  
slength_no_eq/slength_eq

slength_eq/(slength_eq+slength_no_eq)

```

